\Large Inferential Analysis of the Supreme Court Citation Network

### 2018-02-19 Tex/paper.tex
updated title

\documentclass[headsepline=true, abstracton]{scrartcl}
%\usepackage[ngerman]{babel}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{authblk}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{enumerate}
\usepackage{verbatim}
\usepackage[a4paper,text={160mm,215mm},centering,headsep=10mm,footskip=10mm]{geometry}
\usepackage[urlcolor=black,colorlinks=true,linkcolor=black,citecolor=black,bookmarks]{hyperref}
\usepackage{aliascnt}
\usepackage{lmodern}
\usepackage{mdwlist}
\usepackage{natbib}
\usepackage{listings}
\usepackage[table,xcdraw]{xcolor}
\usepackage[activate]{pdfcprot}
\usepackage{graphicx}
\usepackage{slashed}
\usepackage{mathrsfs}
\usepackage{geometry}
\usepackage{float}
\usepackage{oldgerm}
\usepackage{setspace}
\usepackage{dsfont}
%\usepackage{mathtools}
\usepackage[all]{xy}
\usepackage{cite}
%\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{algorithm2e}
\usepackage{url}
\pagestyle{headings}
\newcommand{\myclearpage}{\clearpage}
\newcommand\independent{\protect\mathpalette{\protect\independenT}{\perp}}
\def\independenT#1#2{\mathrel{\rlap{$#1#2$}\mkern2mu{#1#2}}}
\newenvironment{gelaber}{}{}
\newenvironment{preamble}{}{}
\newcommand{\tostar}{\overset{*}{\lower0.5em\hbox{$\smash{\scriptscriptstyle\rightharpoonup}$}}}
\newtheorem{mydef}{Definition}
\newtheorem{bem}{Bemerkung}
\usepackage{tikz}
\newcommand\circlearound[1]{%
  \tikz[baseline]\node[draw,shape=circle,anchor=base] {#1} ;}

@@

\title{\Large Analyzing the Supreme Court Citation Network}
%\author{}
 % Toggle % below to blind and unblind the manuscript
  \author[1]{Christian Schmid \thanks{cxs5700@psu.edu}}
   \author[2]{Ted Hsuan Yun Chen\thanks{thc126@psu.edu}}
 \author[2]{Bruce A. Desmarais\thanks{bdesmarais@psu.edu}}
 \author[1]{David R. Hunter \thanks{dhunter@stat.psu.edu}}
 \affil[2]{Department of Political Science, Pennsylvania State University}
 \affil[1]{Department of Statistics, Pennsylvania State University}

\maketitle
\begin{abstract}
\noindent The significance and influence of US Supreme Court majority opinions derive in large part to opinions' roles as precedents for future opinions. A growing body of literature seeks to understand what drives the use of opinions as precedents through the study of Supreme Court case citation patterns. We raise two limitations of existing work on Supreme Court citaitons. First, dyadic citations are typically aggregated to the case level before they are analyzed. Second, citaitons are treated as if they arise independent of one another. We make the case that studying Supreme Court citations as a dynamic citation network overcomes both of these limitations. We present a methodology for studying citations between US Supreme Court opinions at the dyadic level, as a network. This methodology---the citation temporal exponential random graph model---enables researchers to account for the effects of case characteristics and complex forms of network dependence in citation formation. We apply this methodology to a network that includes all Supreme Court cases decided between 1937 and 2001. We find evidence for dependence processes, including reciprocity, transitivity, popularity, and activity. The dependence effects that we identify are as substantively and statistically significant as the effects of the exogenous covariates we include in the model. The summary result from this analysis is that theoretical models of Supreme Court citation formation should consider both the effects of case characteristics and the structure of past citations.\footnote{
Prepared for presentation at the ``The Enemy of My Enemy: Judicial Politics meets Network Analysis'' conference within a conference at the 2018 Annual Conference of the Southern Political Science Association. This work was supported in part by NSF grants SES- 1558661, SES-1637089, SES-1619644, and CISE-1320219.}
\end{abstract}

@@

There exist two broad reasons why empirical analyses of citations are best defined on the directed dyad level, not the case level. The first is that directed dyadic analyses can test both dyadic and case-level hypotheses. For example, case-level analyses can model whether opinions supported by a liberal majority coalition are more likely than those supported by a conservative majority coalition to be cited heavily in the future, but they cannot precisely model whether liberal cases will be cited more by liberal cases than by conservative cases. Thus, the first reason for analyzing citations at the dyadic level it to expand the set of hypotheses that can be represented in the model. The second reason for studying citations at the directed dyadic level is that, as articulated in the growing literature on legal citation networks, citations form complex networks in which a citation at one point in time may influence future citations. This phenomenon of complex dependence is very common in networks of many types, but processes specific to Supreme Court citations create interdependence in citations. For example, if opinion $i$ relies heavily on opinion $j$ as precedent, opinion $i$ is likely to discuss the legal basis for opinion $j$, and as a consequence cite some of the opinions cited by opinion $j$. Suppose opinion $k$ is cited by opinion $j$. Opinion $k$ is more likely to be cited by opinion $i$ because opinion $i$ relies heavily on $j$, and opinion $j$ cites $k$.  This is a special case of a very common process on networks referred to as ``triad closure''. Complex dependence is theoretically interesting on its own merits, but the effects of covariates cannot be reliably identified---either in terms of coefficient values or standard errors---without accounting for the interdependence inherent in networks \citep{desmaraisstatistical}.

@@

We incorporate a statistic to model the degree to which cases cite those that are similar in terms of the ideological positions of the justices who supported the decision. We account for this effect following citet{spriggs2001explaining}, who find that cases are more likely to be overruled when the Court is ideologically distant from the median justice in the majority coalition that decided the case. \citet{clark2010locating} estimates a latent coordinate model of Cupreme Court opinions based on the network of case-to-case citations. They find that the majority opinion falls at the ideal point of the median member of the majority coalition in the case. We include a covariate term in which $X_{ij}$ is the absolute difference between the Martin-Quinn scores \citep{martin2002dynamic} of the median justices in the majority coalitions for cases $i$ and $j$. We expect this variable to have a negative effect, which would correspond to cases citing those to which they are ideologically similar.%WE COULD USE THIS DATASET AS A SECOND APPLICATION

@@

 \section{Empirical Analysis}
Our three data sources for this study include the SCDB, Martin-Quinn scores, and Supreme Court citation data provided by James Fowler and colleagues \citep{fowler2007network,fowler2008authority}. We limit the Supreme Court terms included in our analysis to those that are covered by all three of these data source (1937--2001).  The Supreme Court citation network from $1937 - 2001$ consists of $8,817$. The breakdown of the data by the Court's Chief Justice is presented in Table \ref{tab:chiefs}. The network has a total of $93,263$ ties. The number of triangles in the network is $211,855$. The in- and outdegree distributions (i.e., the distributions of the number of citations sent and received by cases, respectively), are visualized in figures \ref{indegree_dist} and \ref{outdegree_dist}. The maximum indegree is $190$ and the maximum outdegree is $159$.

\begin{table}[htp]
\centering
\begin{tabular}{|
>{\columncolor[HTML]{C0C0C0}}l |l|l|l|}
\hline
{\color[HTML]{333333} } & \cellcolor[HTML]{C0C0C0}{\color[HTML]{333333} Terms} & \cellcolor[HTML]{C0C0C0}{\color[HTML]{333333} Total Number Cases} & \cellcolor[HTML]{C0C0C0}{\color[HTML]{333333} Cases/Term} \\ \hline
CE Hughes*              & 1937 - 1941                                          & 629                                                               & 125.5                                                     \\ \hline
HF Stone                & 1942 - 1946                                          & 766                                                               & 153.2                                                     \\ \hline
FM Vinson               & 1946 - 1953                                          & 788                                                               & 98.5                                                      \\ \hline
E Warren                & 1954 - 1969                                          & 2159                                                              & 127.0                                                     \\ \hline
WE Burger               & 1970 - 1986                                          & 2805                                                              & 155.8                                                     \\ \hline
W Rehnquist**            & 1987 - 2001                                          & 1670                                                              & 83.5                                                      \\ \hline
\end{tabular}
\caption[caption]{For the time range of interest (1937 - 2001) this table displays the chief justices, the time range they served as chief justice, the number of cases in their time range as well as the average number of cases per year.\\\hspace{\textwidth} * CE Hughes served as chief justice from 1930 - 1941. \\\hspace{\textwidth} ** W Rehnquist served as chief justice from 1987 - 2005.}
\label{tab:chiefs}
\end{table}

>>>>

\documentclass[headsepline=true, abstracton]{scrartcl}
%\usepackage[ngerman]{babel}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
%\usepackage{authblk}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{enumerate}
\usepackage{verbatim}
\usepackage[a4paper,text={160mm,215mm},centering,headsep=10mm,footskip=10mm]{geometry}
\usepackage[urlcolor=black,colorlinks=true,linkcolor=black,citecolor=black,bookmarks]{hyperref}
\usepackage{aliascnt}
\usepackage{lmodern}
\usepackage{mdwlist}
\usepackage{natbib}
\usepackage{listings}
\usepackage[table,xcdraw]{xcolor}
\usepackage[activate]{pdfcprot}
\usepackage{graphicx}
\usepackage{slashed}
\usepackage{mathrsfs}
\usepackage{geometry}
\usepackage{float}
\usepackage{oldgerm}
\usepackage{setspace}
\usepackage{dsfont}
%\usepackage{mathtools}
\usepackage[all]{xy}
\usepackage{cite}
%\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{algorithm2e}
\usepackage{url}
\pagestyle{headings}
\newcommand{\myclearpage}{\clearpage}
\newcommand\independent{\protect\mathpalette{\protect\independenT}{\perp}}
\def\independenT#1#2{\mathrel{\rlap{$#1#2$}\mkern2mu{#1#2}}}
\newenvironment{gelaber}{}{}
\newenvironment{preamble}{}{}
\newcommand{\tostar}{\overset{*}{\lower0.5em\hbox{$\smash{\scriptscriptstyle\rightharpoonup}$}}}
\newtheorem{mydef}{Definition}
\newtheorem{bem}{Bemerkung}
\usepackage{tikz}
\newcommand\circlearound[1]{%
  \tikz[baseline]\node[draw,shape=circle,anchor=base] {#1} ;}

@@

\title{\Large Inferential Analysis of the Supreme Court Citation Network}
%\author{}
 % Toggle % below to blind and unblind the manuscript
  \author[1]{Christian Schmid \thanks{cxs5700@psu.edu}}
   \author[2]{Ted Hsuan Yun Chen\thanks{thc126@psu.edu}}
 \author[2]{Bruce A. Desmarais\thanks{bdesmarais@psu.edu}}
 \author[1]{David R. Hunter \thanks{dhunter@stat.psu.edu}}
% \affil[2]{Department of Political Science, Pennsylvania State University}
 %\affil[1]{Department of Statistics, Pennsylvania State University}

\maketitle
\begin{abstract}
\noindent The significance and influence of US Supreme Court majority opinions derive in large part to opinions' roles as precedents for future opinions. A growing body of literature seeks to understand what drives the use of opinions as precedents through the study of Supreme Court case citation patterns. We raise two limitations of existing work on Supreme Court citations. First, dyadic citations are typically aggregated to the case level before they are analyzed. Second, citaitons are treated as if they arise independent of one another. We make the case that studying Supreme Court citations as a dynamic citation network overcomes both of these limitations. We present a methodology for studying citations between US Supreme Court opinions at the dyadic level, as a network. This methodology---the citation temporal exponential random graph model---enables researchers to account for the effects of case characteristics and complex forms of network dependence in citation formation. We apply this methodology to a network that includes all Supreme Court cases decided between 1937 and 2001. We find evidence for dependence processes, including reciprocity, transitivity, popularity, and activity. The dependence effects that we identify are as substantively and statistically significant as the effects of the exogenous covariates we include in the model. The summary result from this analysis is that theoretical models of Supreme Court citation formation should consider both the effects of case characteristics and the structure of past citations.\footnote{
Prepared for presentation at the ``The Enemy of My Enemy: Judicial Politics meets Network Analysis'' conference within a conference at the 2018 Annual Conference of the Southern Political Science Association. This work was supported in part by NSF grants SES- 1558661, SES-1637089, SES-1619644, and CISE-1320219.}
\end{abstract}

@@

There exist two broad reasons why empirical analyses of citations are best defined on the directed dyad level, not the case level. The first is that directed dyadic analyses can test both dyadic and case-level hypotheses. For example, case-level analyses can model whether opinions supported by a liberal majority coalition are more likely than those supported by a conservative majority coalition to be cited heavily in the future, but they cannot precisely model whether liberal cases will be cited more by liberal cases than by conservative cases. Thus, the first reason for analyzing citations at the dyadic level is to expand the set of hypotheses that can be represented in the model. The second reason for studying citations at the directed dyadic level is that, as articulated in the growing literature on legal citation networks, citations form complex networks in which a citation at one point in time may influence future citations. This phenomenon of complex dependence is very common in networks of many types, but processes specific to Supreme Court citations create interdependence in citations. For example, if opinion $i$ relies heavily on opinion $j$ as precedent, opinion $i$ is likely to discuss the legal basis for opinion $j$, and as a consequence cite some of the opinions cited by opinion $j$. Suppose opinion $k$ is cited by opinion $j$. Opinion $k$ is more likely to be cited by opinion $i$ because opinion $i$ relies heavily on $j$, and opinion $j$ cites $k$.  This is a special case of a very common process on networks referred to as ``triad closure''. Complex dependence is theoretically interesting on its own merits, but the effects of covariates cannot be reliably identified---either in terms of coefficient values or standard errors---without accounting for the interdependence inherent in networks \citep{desmaraisstatistical}.

@@

We incorporate a statistic to model the degree to which cases cite those that are similar in terms of the ideological positions of the justices who supported the decision. We account for this effect following citet{spriggs2001explaining}, who find that cases are more likely to be overruled when the Court is ideologically distant from the median justice in the majority coalition that decided the case. \citet{clark2010locating} estimates a latent coordinate model of Supreme Court opinions based on the network of case-to-case citations. They find that the majority opinion falls at the ideal point of the median member of the majority coalition in the case. We include a covariate term in which $X_{ij}$ is the absolute difference between the Martin-Quinn scores \citep{martin2002dynamic} of the median justices in the majority coalitions for cases $i$ and $j$. We expect this variable to have a negative effect, which would correspond to cases citing those to which they are ideologically similar.%WE COULD USE THIS DATASET AS A SECOND APPLICATION

@@

 \section{Empirical Analysis}
Our three data sources for this study include the SCDB, Martin-Quinn scores, and Supreme Court citation data provided by Courtlistener ?cite. We limit the Supreme Court terms included in our analysis to those that are covered by all three of these data source (1937--2015).  The Supreme Court citation network from $1937 - 2015$ consists of $9,841$ cases. The breakdown of the data by the Court's Chief Justice is presented in Table \ref{tab:chiefs}. The network has a total of $110,742$ ties. The number of triangles in the network is $246,134$. The in- and outdegree distributions (i.e., the distributions of the number of citations sent and received by cases, respectively), are visualized in figures \ref{indegree_dist} and \ref{outdegree_dist}. The maximum indegree is $228$ and the maximum outdegree is $162$.

\begin{table}[htp]
\centering
\begin{tabular}{|
>{\columncolor[HTML]{C0C0C0}}l |l|l|l|}
\hline
{\color[HTML]{333333} } & \cellcolor[HTML]{C0C0C0}{\color[HTML]{333333} Terms} & \cellcolor[HTML]{C0C0C0}{\color[HTML]{333333} Total Number Cases} & \cellcolor[HTML]{C0C0C0}{\color[HTML]{333333} Cases/Term} \\ \hline
CE Hughes*              & 1937 - 1941                                          & 628                                                               & 125.6                                                     \\ \hline
HF Stone                & 1942 - 1946                                          & 756                                                               & 151.2                                                     \\ \hline
FM Vinson               & 1946 - 1953                                          & 786                                                               & 98.25                                                      \\ \hline
E Warren                & 1954 - 1969                                          & 2123                                                              & 124.88                                                     \\ \hline
WE Burger               & 1970 - 1986                                          & 2755                                                              & 153.06                                                     \\ \hline
W Rehnquist            & 1987 - 2005                                          & 2004                                                              & 105.47                                                      \\ \hline
J Roberts **           & 2006 - 2015                                          & 789                                                              & 78.9                                                      \\ \hline
\end{tabular}
\caption[caption]{For the time range of interest (1937 - 2015) this table displays the chief justices, the time range they served as chief justice, the number of cases in their time range as well as the average number of cases per year.\\\hspace{\textwidth} * CE Hughes served as chief justice from 1930 - 1941. \\\hspace{\textwidth} ** J Roberts still serves as chief justice (2/2018).}
\label{tab:chiefs}
\end{table}
